---
title: "Getting Started with cauda"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with cauda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

**cauda** (Causal Automated Unified Data Analysis) is a high-level R toolkit for exploratory causal data analysis. Given any tabular data frame, cauda automatically answers the question:

> *"What variables appear to drive the outcomes in this dataset, and how do the outcomes depend on those drivers?"*

The package combines automated causal discovery, interactive graph editing, correlation analysis, machine learning-based visualizations, and decision optimization into a single consistent interface.

---

## Installation

```{r install, eval = FALSE}
# Install from local source
devtools::install("path/to/cauda R")

# Load the package
library(cauda)
```

---

## Quick Start

The fastest way to get a full analysis is with a single command:

```{r quickstart}
library(cauda)

# Load any data frame
df <- read.csv("your_data.csv", stringsAsFactors = FALSE)

# Run the full pipeline
results <- cauda.analyze(df, highlight = "YourTarget")
```

This automatically runs missing value checks, recodes and cleans the data, learns a causal network, and produces correlation visualizations — all in one shot.

---

## Step-by-Step Walkthrough

We demonstrate cauda using the **TelcoChurn** dataset, which contains customer information for a telecom company. The goal is to understand what drives customer churn.

### Step 1 — Check for missing values

Before doing anything, inspect the raw data:

```{r missing}
df <- read.csv("TelcoChurn.csv", stringsAsFactors = FALSE)
cauda.missing(df)
```

This prints a table showing how many missing values each variable has and flags any variables above the 5% threshold.

### Step 2 — Prep the data

`cauda.prep()` handles all data wrangling automatically:

```{r prep}
df_ready <- cauda.prep(df)
```

Under the hood this:

- Recodes character variables (Yes/No → 1/0, Male/Female → 1/0, multi-level factors → integers)
- Drops high-cardinality columns like customer IDs
- Removes near-zero variance columns
- Drops highly correlated redundant columns
- Handles missing values

### Step 3 — Learn a causal network

```{r dag}
dag <- cauda.dag(df_ready, highlight = "Churn")
```

This uses the Hill-Climbing algorithm from `bnlearn` to automatically discover a directed acyclic graph (DAG) showing dependency relationships between variables. The target variable (`Churn`) is highlighted in red.

### Step 4 — Edit the network

The automatically discovered DAG can be refined using domain knowledge:

```{r edit}
# Remove an arrow that doesn't make causal sense
dag <- cauda.delete(dag, "Churn", "VoiceMailPlan", highlight = "Churn")

# Flip an arrow direction
dag <- cauda.flip(dag, "Churn", "InternationalPlan", highlight = "Churn")

# Add a new arrow
dag <- cauda.add(dag, "tenure", "TotalRevenue", highlight = "Churn")
```

Each command replots the DAG automatically so you can see the change.

### Step 5 — Correlation analysis

View pairwise and partial correlations side by side:

```{r corr}
# Full correlation heatmap + spring network
cauda.corr(df_ready, highlight = "Churn")

# Partial correlation network (controls for other variables)
cauda.pcorr(df_ready, highlight = "Churn")
```

Partial correlations are particularly useful — they show the relationship between two variables *after* removing the influence of all other variables, which is much closer to actual causal signal.

### Step 6 — Consensus causal network

Run multiple causal discovery algorithms and see where they agree:

```{r consensus}
cauda.consensus(df_ready, highlight = "Churn")
```

Thicker, darker arrows indicate more algorithms agreed on that dependency. This gives a sense of which relationships are robust versus which are algorithm-specific.

### Step 7 — Visualize feature effects

Use partial dependence plots to see how individual features affect the outcome:

```{r pdp}
# How does tenure affect churn probability?
cauda.pdp(df_ready, target = "Churn", feature = "tenure", method = "pdp")

# ICE plot — shows individual-level variation
cauda.pdp(df_ready, target = "Churn", feature = "tenure", method = "ice")

# 2D PDP — how do tenure and Contract interact?
cauda.pdp2d(df_ready, target = "Churn", feature1 = "tenure", feature2 = "Contract")
```

### Step 8 — Conditional independence tests

Test which variables are genuinely dependent on the target after controlling for everything else:

```{r independence}
cauda.independence(df_ready, target = "Churn")
```

A low p-value means the variable is genuinely associated with Churn even after controlling for all other variables.

### Step 9 — Decision optimization

Find the optimal settings for controllable variables to minimize or maximize an outcome:

```{r optimize}
cauda.optimize(
  df_ready,
  target     = "Churn",
  controls   = c("tenure", "Contract", "TechSupport"),
  actionable = c("Contract", "TechSupport"),  # tenure is not actionable
  maximize   = FALSE
)
```

The `actionable` argument distinguishes variables you can actually intervene on (like contract type) from descriptive variables you cannot control (like how long someone has been a customer).

---

## Key Findings on TelcoChurn

Running the full cauda pipeline on TelcoChurn reveals several actionable insights:

- **Contract type** is the strongest protective factor against churn — customers on two-year contracts churn at dramatically lower rates than month-to-month customers
- **Tech support** and **online security** significantly reduce churn when present
- **Tenure** is strongly negatively correlated with churn — newer customers are the highest risk group
- **International plan** and **customer service calls** are the strongest positive predictors of churn
- **Gender** is completely isolated in the causal network — it has no dependency relationship with any other variable

---

## Function Reference

| Function | Description |
|----------|-------------|
| `cauda.missing(df)` | Summarize missing values |
| `cauda.recode(df)` | Recode character variables to numeric |
| `cauda.clean(df)` | Remove uninformative/redundant columns |
| `cauda.prep(df)` | Recode + clean in one step |
| `cauda.dag(df, highlight)` | Learn and plot causal DAG |
| `cauda.add(dag, x, y)` | Add arrow to DAG |
| `cauda.delete(dag, x, y)` | Delete arrow from DAG |
| `cauda.flip(dag, x, y)` | Flip arrow direction in DAG |
| `cauda.corr(df, highlight)` | Correlation heatmap + network |
| `cauda.pcorr(df, highlight)` | Partial correlation network |
| `cauda.consensus(df, highlight)` | Consensus DAG across algorithms |
| `cauda.pdp(df, target, feature)` | PDP / ICE / ALE plots |
| `cauda.pdp2d(df, target, f1, f2)` | 2D partial dependence plot |
| `cauda.optimize(df, target, controls)` | Decision optimization |
| `cauda.independence(df, target)` | Conditional independence tests |
| `cauda.analyze(df, highlight)` | Full pipeline in one command |
